---
title: "Project Proposal"
author: "Patricia Gribi"
date: "2023-10-02"
output: html_document
---


# Questions:

Objective: goal of the project. What will be the result, answering what question?
- What do you think of my ideas?
- Do I need one specific research question? And it can be something that already exists, right?


Data
- Should I know already which datasets I want to use exactly for the proposal?
- Should I consider also other datasources than the ones suggested? Do you know if the suggested data is good data?
- How many different sources?
- If I do not have complete data model data?


Analysis
- What if I realize during the process, that my analyises are too hard?


Proposal
- Cite studies in the proposal? (what is known, what unknown.) And what about later on?
- for the timeline: I do not really now yet what exactly I will have to do, are the following steps enough?
- Implementation: I do not now already
- Impact: How will your results answer unknowns? -- Does my results have to answer unknowns?


# Summary

This research project aims to reconstruct a time series of global mean CO2 and temperature data, from the Pleistocene to 2100. In a first step different data used for global mean CO2 concentration and temperature records will be identified and described.The sources, methodologies and associated challenges will be discussed. In a further step the time series will be analyzed. This includes historical trends, such as the industrial revolution, and making projections about the future. The latter considering different emission and temperature scenarios.

Objective Ideas: reconstruct time series of co2 and temp., what is anthropogenic and what is the natural variability of climate, analyze historical trends,
projections about future, compare northern and southern hemisphere, compare land vs. ocean, outlier years...


# Background and Motivation

Why is your project important? What is known and unknown?

Why important: current climate debate, climate change, future scenarios,  

this is known: glacial and interglacial periods, co2 amount in the atmosphere varies, in the last
century there has been a strong increase in co2 levels due to anthropogenic activities, 
the amount of co2 in the atmosphere correlates with the temperature -- the average global temperature
is rising. 

this is unkown: overall uncertainty of paleo-based estimates of climate sensitivity


# Objective

Describe the goal of your project. What will be the result, answering what question?

- Homogenized time series of global mean CO2 and temperature, from the Pleistocene to 2100.
datasets: Mauna Loa CO2 trends, Corrected EPICA Dome C CO2 record (Bereiter et al. 2015), RCP standard forcing


# Implementation

What approach will you take to achieve the goal in terms of data and methods?


# Papers

- Pleistocene sea surface temp. reconstructions: https://www.sciencedirect.com/science/article/abs/pii/S0012821X1930603X
Our uncertainty analysis reveals further that the lack of robust reconstructions of glacial aerosol forcing is a key contributor to the overall uncertainty of paleo-based estimates of climate sensitivity.The global-mean temperature reconstruction is typically based on an average of local ocean and/or land temperatures derived from different paleo-proxies. The accuracy of these “paleothermometers” is uncertain.
reconstructions very difficult because proxies can be local or influenced by other parameters, models on the other hand
cannot take into consideration the earth system as a whole and complexity is therefore largely restricted (letzter Abschnitt introduction). 

radiative forcing: change in energy balance in the Earth's atmosphere due to various factors that influence the Earth's climate. It quantifies the impact of different drivers on the Earth's energy budget, specifically in terms of the incoming and outgoing energy fluxes.primary factors, that can cause radiative forcing: greenhouse gases, aerosols, land use changes, solar variability, stratospheric water vapor

SST: sea surface temperature


- Identifying Signatures of Natural Climate Variability in Time Series of Global-Mean Surface Temperature:distinction between anthropogenic influences and natural variability of the climate (vulcanic eruptions, el niño, marine air masses over high-latitude continents in winter)

- Measuring Global Warming: very nice figures and well explained, FTS model functional time series model, obtain forecasts for next twenty years

"temperature anomalies": differences btw. observed temp reading and the long-term average temp for each location and time


# Datasources

- Mauna Loa CO2 trends: Globally averaged marine surface annual mean data, from 1979 until 2022, in ppm
- Corrected EPICA Dome CO2 record:
- RCP standard forcing: Representative Concentration Pathways (RCPs), different regions and scenarios, https://tntcat.iiasa.ac.at/RcpDb/dsd?Action=htmlpage&page=download , 

What about temperature data?




# Responsibilities and timeline

Session 6 (25.10.) : Final proposal and Data Research
Session 7 (01.11.) : Tidy data
Session 8 (08.11.) : Visualize data
Session 9 (15.11.) : Visualize data
Session 10 (22.11) : Analysis
Session 11 (29.11) : Analysis
Session 12 (06.12) : Writing report, Interpretation and Conclusion
Session 13 (13.12) : Create Presentation, Finish report
Session 14 (20.12) : Presentations. Report-Abgabe.


# Risks and contingency

What risks and challenges do you expect and what will you do to address them?

Expected risks: incomplete data, the whole data wrangling and data cleaning process
to have tidy data in the end, as the data come from different sources, with diff. 
variables, layouts..., 

challenges: to know how much to go deep with my analyses, to actually find out what
exactly to do


# Impact
How will your results answer unknowns? What are the expected scientific and societal impacts?
- Expected scientific impact(s): e.g. contributing to specific scientific advances, across and within disciplines, reinforcing research infrastructures (databases, equipment, and instruments)
- Expected societal impact(s): e.g. improving policies and decision-making, providing guidance for sustainable management, raising consumer awareness.
Only include such impacts where your project would make a direct contribution.







# Data Exploration

Explore the selected datasets to get a sense of the data's structure, variables, and quality. Check for any missing or erroneous data that may need to be cleaned.


# Data Cleaning and Preprocessing

Prepare the data for analysis by cleaning and transforming it as needed. This may include handling missing values, standardizing units, and merging datasets if multiple data sources are used.


# Data Analysis and Visualization

Apply statistical and data visualization techniques to analyze the data. Use appropriate tools and libraries (e.g., Python with libraries like pandas, matplotlib, and seaborn) to create visualizations that reveal patterns, relationships, trends, or differences.


# Hypothesis Testing (if applicable):

If your research question involves hypothesis testing, perform the necessary statistical tests to validate or reject your hypotheses.


# Interpretation and Conclusion:

Interpret the results of your analysis in the context of your research question. Discuss the implications and significance of your findings. Address what is known and what remains unknown in the field.



```{r}
terra::rast(paste0(here::here(), "/data/testdata/demo_data.nc"))
```


```{r}
install.packages("raster")
library(raster)
```



```{r}

current_directory <- getwd()
print(current_directory)


library(terra)

raster_data <- terra::rast("./demo_data.nc")

```

